\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-19 paper.qmd
Added total hex bins with new hex-binning approach

@@

```{r}
#| eval: false
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** training data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. What is the number of bins that would be effective in representing low-dimensional data?"

@@

umap_data_with_hb_id_pbmc <- as.data.frame(do.call(cbind, hb_obj_s_curve$data_hb_id))

>>>>

```{r}
#| echo: false
#| message: false
#| warning: false

num_bins_x <- 3
num_bins_y <- 4
hex_size <- 0.62

## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj$centroids))
counts_df <- as.data.frame(do.call(cbind, hb_obj$std_cts))

hex_full_count_df_loop1 <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))

p1 <-  ggplot(data = hex_full_count_df_loop1, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_s_curve_scaled, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_equal() +
  #coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| echo: false
#| message: false
#| warning: false

num_bins_x <- 6
num_bins_y <- 14
hex_size <- 0.13

## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj$centroids))
counts_df <- as.data.frame(do.call(cbind, hb_obj$std_cts))

hex_full_count_df_loop2 <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))

p2 <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_s_curve_scaled, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_equal() +
  #coord_cartesian(xlim =c(-1, 2.7), ylim = c(-2, 12.2)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| echo: false
#| message: false
#| warning: false

num_bins_x <- 12
num_bins_y <- 28
hex_size <- 0.06

## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj$centroids))
counts_df <- as.data.frame(do.call(cbind, hb_obj$std_cts))

hex_full_count_df_loop3 <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))

p3 <-  ggplot(data = hex_full_count_df_loop3, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_s_curve_scaled, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  #coord_equal() +
  #coord_cartesian(xlim =c(-1, 2.7), ylim = c(-2, 12.2)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "c", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** embeddings of **S-curve** training data: (a) b = 12 (3, 4), h = 0.62, (b) b = 84 (6, 14), h = 0.13, and (c) b = 336 (12, 28), h = 0.06. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. What is the number of bins that would be effective in representing low-dimensional data?"

@@

```{r}
#| eval: false
#| echo: false
#| label: fig-binsize1
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** training data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. What is the number of bins that would be effective in representing low-dimensional data?"

@@

umap_data_with_hb_id <- as.data.frame(do.call(cbind, hb_obj_s_curve$data_hb_id))

### 2024-03-19 paper.tex
Added total hex bins with new hex-binning approach

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the 2D model. This is because it is the
initial step in building the 2D model. The hexagonal grid with the
chosen total number of bins must be able to capture the structure of the
NLDR data. If the number of bins is too low, the model may not be able
to capture the structure of the NLDR data effectively (see
\textbf{?@fig-binsize} (a)), while if there are too many bins, it may
result in over-fitting the individual points of the NLDR data (see
\textbf{?@fig-binsize} (c)). Therefore, it is important to determine an
appropriate number of bins to build an effective model.

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-87-1.pdf}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

The number of hexagonal bins in the hexagonal grid has a considerable
impact on the construction of the 2D model. This is because it is the
initial step in building the 2D model. The hexagonal grid with the
chosen total number of bins must be able to capture the structure of the
NLDR data. If the number of bins is too low, the model may not be able
to capture the structure of the NLDR data effectively (see
Figure~\ref{fig-binsize} (a)), while if there are too many bins, it may
result in over-fitting the individual points of the NLDR data (see
Figure~\ref{fig-binsize} (c)). Therefore, it is important to determine
an appropriate number of bins to build an effective model.

{\centering \includegraphics{paper_files/figure-pdf/fig-binsize-1.pdf}

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the \textbf{UMAP} embeddings of \textbf{S-curve} training data: (a)
b = 12 (3, 4), h = 0.62, (b) b = 84 (6, 14), h = 0.13, and (c) b = 336
(12, 28), h = 0.06. The hexbins are colored based on the density of
points, with darker colors indicating higher point density and yellow
color representing lower point density within each bin. What is the
number of bins that would be effective in representing low-dimensional
data?}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-91-1.pdf}
